---
title: "Thermal niches of scleractinian corals in deep time - an example application for the chronosphere in R"
author: "Ádám T. Kocsis, Paulina Nätscher & Nussaïbah B. Raja"
date: "`r format(Sys.time(),'%Y-%m-%d')`"
output: 
  pdf_document: default
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(knitr)
opts_knit$set(root.dir=normalizePath('../'),#set directory
              tidy.opts=list(width.cutoff=80), tidy=TRUE #wrap long lines            
) 
```

# 1. Setting up the environment to work with the data

## 1.1. R extension packages 

This example uses the 'chronosphere' R package to download data, and the 'divDyn' R package (Kocsis et al., 2019) to process and plot the results.

```{r libraries and functions, message=FALSE, warning=FALSE}
# libraries
library(chronosphere)
library(divDyn)
```

You can use the `datasets()` function to see which datasets are available *publicly* in the chronosphere library.

```{r datasets}
# datasets table from chronosphere
allDat <- datasets()

head(allDat[,1:3])
```

The loading functions downloaded from the 'chronosphere' servers might need some additional packages. 
These do not have to be attached to the workspace, but they must be present on the system, or the download call will result in an error message, pointing to which package you are missing. 
The spatial temperature data used in this example are dependent on the 'ncdf4' package, which includes methods that read in netcdf binaries as rasters. 
If you do not have this package on your system, you can install it from the CRAN with: 

```{r ncdf, echo=TRUE, eval=FALSE}
install.packages("ncdf4")
``` 


## 1.2. Analytical functions

Data transformation and analytical functions will be shown here, but printing all procedures in this document would make it very technical and cluttered. 
All functions used in this example can be found on its dedicated GitHub repository (`https://github.com/chronosphere-portal/case_coralTherm`), and can be loaded with these simple calls that download and read them to the R workspace. 
If you are interested in what a particular function is doing, type its name into the console.

```{r funct}
# functions to clean, transform and manipulate PBDB data
source("https://github.com/chronosphere-portal/case_coralTherm/raw/master/scripts/methods/pbdb.R")

# functions that calculate new variables using the data
source("https://github.com/chronosphere-portal/case_coralTherm/raw/master/scripts/methods/analysis.R")

# functions that plot results
source("https://github.com/chronosphere-portal/case_coralTherm/raw/master/scripts/methods/plots.R")
```

```{r functionsielnt, echo=FALSE}
# function printing where necessary
source("https://github.com/chronosphere-portal/case_coralTherm/raw/master/scripts/methods/utils.R")
```

# 2. Data download and processing

## 2.1. The Paleobiology Database (PBDB)

The following code downloads PBDB data directly from the `chronosphere` servers. 
The most up-to-date version of these datasets are downloaded by default. 
The default variable (`var="occ"`) covers all data in the Ediacaran-Holocene interval. 

```{r get_pbdb, message=FALSE}
pbdb <- fetch(dat="pbdb") #PBDB data
```

The backend of 'chronosphere' will try to get a daily copy of the PBDB every day. 
Upon download, you are indicated to acknowledge the database in your publication if you use the data. 
Objects downloaded from the 'chronosphere' get some basic information appended to them as attributes. 
These can be used for get further information, such as the exact version of the download (`ver`) the access date, or the API call used to get the download from the database (`API`). 

```{r pbdb_att, message=FALSE}
# list attributes
names(attributes(pbdb)$chronosphere)

# date of access with API
attributes(pbdb)$chronosphere$accessDate
```

You can set the `datadir` argument of the `fetch()` function to a valid directory, while would save the downloaded data in a specific directory on your computer. 
These can can be loaded again in the future without an internet connection and it also ensures that the same data are used everytime the analyses are run.

In this example, we will be working with occurrrences data of scleractinian corals.

```{r corals}
#subset for scleractinian corals only
corals <- pbdb[pbdb$order =="Scleractinia",] #only scleractinia
```

## 2.2. Traits of fossil scleractinian genera

These raw occurrences can be processed further merged and merged trait information which was used earlier in Kiessling and Kocsis (2015), about the diversity dynamics of coral genera. 
The raw data-table used for this analysis is added to the the supplementary data-subset of the chronosphere. 
Unlike other big databases, this (`dat="som"`) contains a variety of smaller datasets (variables) that were published as Supplementary Online Material. 

```{r get_traits, message=FALSE}
traits <- fetch(dat="som", var="kiessling-coralgenera") #coral traits data
```

This `data.frame` objects indicate how occurrence names in the PBDB ('genus_detail') match with a validated lists of genus names ('genus_proper'). 
We can use this list to merge the pbdb with trait data, and treat the two symbiotic groups separately. 

```{r coralsmerge}
# assign the symbiotic status
corals <- merge(
  corals, 
  unique(traits[, c("genus_detail", "genus.proper", "ECOLOGY")]), 
  by.x="genus", by.y="genus_detail", all.x=TRUE)

```

## 2.3. Stratigraphic assignment and further cleaning

Before we can analyses the data, we will bin the data at stage level using our custom function `AssignStages()`:
```{r assign_stages, echo=FALSE, results='asis'}
printfunction("AssignStages")
```

Note that this function is based on the manual accompanying the R package divDyn ($divDyn), and this snippet cannot assign occurrences from the Cambrian and Ordovician periods. 
As scleractinian corals do not have Early Paleozoic record, they are not affected by this issue. 

```{r}
# Assign stage to PBDB entries
corals <- AssignStages(corals)
```

We also remove all occurrences without any genus names and coordinates, as well as those that could not be assigned to a specific stage. 
Additionally, we subset the data to include coral occurrences only from the Triassic onwards. 
This is done using our custom function `CleanPBDB`:

```{r clean_pbdb, echo=FALSE, results='asis'}
printfunction("CleanPBDB")
```

```{r}
# Assign stage to PBDB entries
corals <- CleanPBDB(corals)
```

We also omit collections for which no coordinates were available or those that could not be assigned a stage. 
```{r}
# separate the collection data
collections <- corals[,c("collection_no", "lng", "lat", "stg", "paleolat" ,"paleolng")]
collections <- unique(collections)

# omit cases where something is missing
collections <- collections[
  !is.na(collections$stg) & 
  !is.na(collections$lng) & 
  !is.na(collections$lat), ]

```

## 2.4. Sea Surface Temperature (SST) data 

For our analyses, we also need sea surface temperature (SSTs) data. 
The following were computed by Valdes et al. (2017) using the HadCML3 climate models. 
These were spatially interpolated to best match Paleobiology Database occurrence data (will be expanded later).

These are not yet added to the public library. 
To reach these items, we have to connect to the private, TERSANE chronosphere library. 
This can be achieved with the function below, that adjusts the address of the remote server.

```{r useall, echo=FALSE, results='asis'}
source("../misc/useall.R")
printUseall("../misc/useall.R")
```

```{r sst, warning=FALSE, message=FALSE}
useall()
tos <- fetch(dat="had-stage", var="tos-extra", res=1)
```

## 2.5. Paleocoordinate reconstructions  

Paleocoordinate reconstruction can be accomp accomplished in two way with the 'chronosphere' R package.
We also need the rotation model for the coordinate reconstructions. **Note:** This example uses the offline reconstruction feature for Gplates. 

```{r paleomap, message=FALSE}
# download the rotation model
mod <- fetch(dat="paleomap",var="model")

```

To be able to use our SST data together with our genus occurrences, we will match the closest temperature layer available to each stage in our data. This is important as we have missing data for some stages and hence this allows us to get an approximate temperature reconstruction for those stages with missing data.This can be done using the `matchtime` function:

```{r}
# Order the TOS series
tosord <- matchtime(tos, stages$mid)

# assign a reconstruction to every collection
collections$mapage <- names(tosord)[collections$stg]

```

We also reconstruct the coordinates of our genus occurrences using our specified model. While the PBDB already provide reconstructed paleocoordinates, we need to match the model used for reconstruction of the paleocoordinates to the temperature data to allow for maximum accuracy. 

```{r}
# Offline reconstruction - might take a minute or two
reconstructed <- reconstruct(collections[, c("lng", "lat")], 
                             age=collections[, "mapage"], enumerate=FALSE, model = mod)

# bind the new coordinates to the previous
colnames(reconstructed) <- c("plng", "plat")
collections <- cbind(collections, reconstructed)

```

We can finally extra SST values for each collections. We also omit those occurrences without a corresponding SST value.

```{r}
# extract the SST values
collections$SST <- chronosphere::extract(tosord, collections, by="mapage")

# omit everything with no SST data
collSST <- collections[!is.na(collections$SST), ]

```

```{r plot_sst, fig.width=10, fig.asp=0.67}
plot(0,0, xlim=c(-180,180), ylim=c(-90,90), type="n", xaxs="i", yaxs="i")
mapplot(tos["130"], add=TRUE)
points(collSST[collSST$mapage==130, c("plng", "plat")],pch=16, col="black")
```

We then add the SST values to our genus occurrences:
```{r}
# remerge with occurrences
corals <- merge(
  corals, 
  collSST[, c("collection_no", "plat", "plng", "SST", "mapage")], 
  by="collection_no")

```

# Thermal niches of scleractinian corals

Now that our data is in the right format, we can calculate the thermal niche of the zoothanthellate and azoothanthellate over time.

```{r}
# compare Z with AZ corals
zCorals <- corals[corals$ECOLOGY=="z", ]
azCorals <- corals[corals$ECOLOGY=="az" | corals$ECOLOGY=="ap", ]
```

Spatial distribution can be plotted with this function:


Ranges of zoothanthellate and azoothanthellate corals for the `r sprintf("%s (%g - %g Ma)", stages$stage[75], round(stages$top[75], 1), round(stages$bottom[75], 1))` stage:
```{r, fig.width=12, fig.asp=0.5}
PlotMapWithRanges(zCorals, azCorals, i=75,zcol="#cc2f04", azcol="#0447cc", alpha=60)
```

Ranges of zoothanthellate and azoothanthellate corals for the `r sprintf("%s (%g - %g Ma)", stages$stage[93], round(stages$top[93], 1), round(stages$bottom[93], 1))` stage:
```{r, fig.width=12, fig.asp=0.5}
PlotMapWithRanges(zCorals, azCorals, i=93,zcol="#cc2f04", azcol="#0447cc", alpha=60)

```

Comparisons between the zoothanthellate and azoothanthellate corals show a significant difference between the two groups, with z-corals having a smaller temperature range than the az-corals. With this plotting function


```{r, fig.width=6, fig.asp=0.9}
WT <- wilcox.test(zCorals$SST, azCorals$SST)

# function plots boxplots of z- and az-corals with the underlying distribution
CompareSST(zSST=zCorals$SST, azSST= azCorals$SST,
           col = pal, wilcox=WT)

```

We can tabulate the median genus niche of a coral group with the following calculations
```{r, results='asis', echo=FALSE}
printfunction("GenusNiche")
```



```{r}
# Genus median niche
allMedZ <- GenusNiche(zCorals)
allMedAZ <- GenusNiche(azCorals)
```

Which then can be plotted over time, using a standard geological time plot



```{r}

# time-specific occupied niche - heavilly affected by sampling
# Plot for the SOM
Tp(ylim=c(0,40),ylab=c("Sea Surface Temperature (C)"))
LinesNiche(allMedZ, col=paste0(pal["z"], 55))
LinesNiche(allMedAZ, col=paste0(pal["az"], 55))
```

Assuming niche conservatism, i.e. that the individual genus did not change its thermal niche over time, we calculate the median thermal niche per genera over time. This can be calculated with the following function
```{r, results='asis', echo=FALSE}
printfunction("GenusNicheOverTime")
```



```{r}
# Calculate the median niche of the genus with its FAD and LAD
flZ <- GenusNicheOverTime(zCorals)
flAZ <- GenusNicheOverTime(azCorals)

Tp(ylim=c(0,40), ylab=expression("Sea Surface Temperature ("*~degree*C*")"))
segments(
  x0=stages$mid[flZ$FAD], 
  x1=stages$mid[flZ$LAD], 
  y0=flZ$var, 
  y1=flZ$var, 
  col=paste0(pal["z"], 22), 
  lwd=2)
segments(
  x0=stages$mid[flAZ$FAD], 
  x1=stages$mid[flAZ$LAD], 
  y0=flAZ$var,
  y1=flAZ$var, 
  col=paste0(pal["az"], 22),
  lwd=2)


meanZ<- MeanOverTime(flZ)
meanAZ <- MeanOverTime(flAZ)

lines(stages[names(meanZ), "mid"], meanZ, col=pal["z"], lwd=5)
lines(stages[names(meanAZ), "mid"], meanAZ, col=pal["az"], lwd=5)


```

We see a decrease in median thermal niche is observed since the Triassic, which could be the result of sampling. Interestingly, a divergence in thermal niche is also seen from the early to mid Cretaceous. 

# References

Kiessling and Kocsiss

Kocsis et al.

Valdes et al. 